---
title: "Week 2 R Notebook Work"
output: html_notebook
---

This is the code from chapters 2-4 of Islam's "Hands-On Geospatial Analysis with R and QGIS"


Start of Chapter 2 Material
```{r}

#Plotting Point Data

# SpatialPoints
library(sp)
library(rgdal)
library(maptools)
map = readOGR("C:/Program Files/RStudio/Hands-On-Geospatial-Analysis-with-R-and-QGIS-master/Chapter02/Data","indicator") # Imports shapefile into R, gives it the name "map". Shapefile cannot have .shp extension.

plot(map)  # Plots the imported shapefile. 

class(map) # Displays class data for "map", which should be SpatialPointsDataFrame

bd_val = read.csv("C:/Program Files/RStudio/Hands-On-Geospatial-Analysis-with-R-and-QGIS-master/Chapter02/Data/r_val.csv", stringsAsFactors = FALSE) # imports shapefile and assigns name "bd_val"

str(bd_val)   # checks and displays structure of dataset

coordinates(bd_val)=c("lon", "lat")  
# Coverts from a standard dataframe to a SpatialPointsDataFrame, same as "map"
str(bd_val)

plot(bd_val, col = "blue", pch = 19) #plots bd_val as blue dots
```

```{r}
#SpatialLines

highway = readOGR("C:/Program Files/RStudio/Hands-On-Geospatial-Analysis-with-R-and-QGIS-master/Chapter02/Data","dhaka_gazipur")
#imports the file, same as before

plot(highway) # creates and displays a map of highways in Dhaka, Bangladesh



map_dhaka = readOGR("C:/Program Files/RStudio/Hands-On-Geospatial-Analysis-with-R-and-QGIS-master/Chapter02/Data","dhaka")


plot(map_dhaka)


str(map_dhaka, max.level = 2) # displays a laconic structure

map_bd = readOGR("C:/Program Files/RStudio/Hands-On-Geospatial-Analysis-with-R-and-QGIS-master/Chapter02/Data","BGD_adm3_data_re")
head(map_bd@data)

str(map_bd@polygons, max.level = 2)

sixth_element = map_bd@polygons[[6]] #activates the 6th polygon in "map_bd"
str(sixth_element, max.level = 2) #succinctly investigates 6th polygon


str(sixth_element@Polygons[[2]], max.level = 2) #checks structure of second polygon within sixth_element
plot(sixth_element@Polygons[[2]]@coords) #plots second polygon's coordinates


map_bd$NAME_3 # Accesses and prints values from the column NAME_3 in map_bd
```
This chunk plots a map with both polygons and points, and contains code for reprojection of map
```{r}

# Points and Polygon Data in One Map

plot(map_bd)
points(bd_val, pch=19, col="blue") 



map_bd = spTransform(map_bd, CRS("+proj=longlat +datum=WGS84")) 

# Reprojects map to WGS84 and longitude/latitude
```

The following chunk produces a choropleth from the dhaka_div shapefile.
```{r}  

#making a chloropleth
library(GISTools)
choropleth(map_bd, as.numeric(map_bd$value2)) #creates choropleth from quantitative values in map_Bd
library(RColorBrewer) #loads RColorBrewer Library
dhaka_div = readOGR("C:/Program Files/RStudio/Hands-On-Geospatial-Analysis-with-R-and-QGIS-master/Chapter02/Data","dhaka_div")

unique(dhaka_div$NAME_3) #identifies number of unique elements within dhaka_div


# Assigning a color ramp to choropleth based on unique values
colors = colorRampPalette(brewer.pal(12,"Set3"))(7)
dhaka_div$NAME_3 = as.factor(as.character(dhaka_div$NAME_3))
spplot(dhaka_div,"NAME_3", main = "Coloring different districts of Dhaka division", col.regions = colors, col = "white")


                              
```
The following Code Chunk will be create several maps of Bangladesh with different map features

```{r}

#install.packages("tmap")   Turned this into a comment after initial install


library(tmap)

# Loads required library used later in chunk



str(map_bd@data)

map_bd$value1 = as.numeric((map_bd$value1))

qtm(shp = map_bd, fill = "value1")

# I lost the plot around this point, and for some reason the values outputted are like 100x greater



tm_shape(map_bd) + 
  tm_borders() +
  tm_fill(col="value1") +
  tm_compass() +
  tmap_style("cobalt")
#adds map features to map (borders for polygons, fill for polygons, background, compass)

tm_shape(map_bd) +
 tm_bubbles(size = "value1", style = "quantile") +
 tm_borders(col="orange3")
#Makes a map with orange border around polygons, does not fill polygons, adds dots with scaling based on value1

tm_shape(map_bd) +
 tm_fill(col = "value1", style = "quantile") +
 tm_borders() +
 tm_text(text = "NAME_3", size = 0.5)
# Labels map with 0.5 size labels to the map created in the map created between lines 119 and 123
  

tm_shape(map_bd) +
 tm_fill(col = "value1", style = "quantile", title = "Value of quantitative
indicator", palette = "Blues") +
 tm_borders(col = "grey30", lwd = 0.6) +
 tm_text(text = "NAME_3", size = 0.5) +
 tm_credits("Source: Author", position = c("right", "top"))
# Adds author name in top right, fills polygon with a different color ramp (blue, not green) adds title to legend, adds labels to polygons

```


End of Chapter 2 Material

Start of Chapter 4 Material

Combining Shapefiles
```{r}


library(sp)
library(rgdal)
library(maptools)

# Loads Libraries required for following code


isGazipur = map_bd$NAME_3 == "Gazipur"
# Creates a variable (not sure if this is the correct term, that's what we'd call it in Matlab) that encompasses those features in map_bd with the name value of "Gazipur"


gazipur = map_bd[isGazipur, ]



# writeOGR(obj=gazipur, dsn="C:/Program Files/RStudio/Hands-On-Geospatial-Analysis-with-R-and-QGIS-master/Chapter04/Data", layer="gazipur", driver="ESRI Shapefile")

# Saves the map as a shapefile named gazipur (though throws an error after first use, so will likely make this a comment before running) using the writeOGR() function


plot(gazipur, col = "blue")

#Displays map in blue

dhaka = readOGR("C:/Program Files/RStudio/Hands-On-Geospatial-Analysis-with-R-and-QGIS-master/Chapter02/Data","dhaka")

dhaka_gazipur = raster::union(dhaka, gazipur)
str(dhaka_gazipur, max.level = 2)
plot(dhaka_gazipur, col = "blue")

# The important line here is 187, as that is merging the maps of dhaka and gazipur into a single shapefile. The rest is stuff that has appeared before in this notebook thing.




```











Learning how to do Clipping in R
```{r}

points = read.csv("C:/Program Files/RStudio/Hands-On-Geospatial-Analysis-with-R-and-QGIS-master/Chapter04/Data/arbitrary_indicator.csv")

# assigns the csv file of interest to the object 'points'. Don't put the quotations outside of read.csv, or you'll spend 30 minutes troubleshooting the coordinates command when the read.csv just never got run.

coordinates(points) = ~ lon + lat

# allows the data in the csv file to be read as coordinates, which is used shortly to plot the points as dots on a merged map map.

summary(points)
#provides a summary of data in object "points"

plot(map_bd, col = "gray", border = "blue", main = "Map of Bangladesh with
arbitrary points plotted")
plot(points, add = TRUE, pch=19, cex=.4, col = "red")

#plots the map of bangladesh with the points in "points" as a single image. The points are displayed as red crosses, while the bangladesh map is displayed with a grey fill and blue border. The map is given the title "Map of Bangladesh with Arbitrary Points Plotted.

proj4string(points) <- CRS(proj4string(dhaka))
points_dhaka <- points[dhaka, ]

# sets the projection of points to match that of dhaka, then clips the points to the map of dhaka 

plot(dhaka, col = "gray", border = "blue", main = "Points clipped to the
map of Dhaka")
plot(points_dhaka, add = TRUE, pch=20, cex=1, col = "red")




not_dhaka = gDifference(points, dhaka)
plot(map_bd, col = "gray", border = "blue", main = "All the points except
Dhaka's are plotted")
plot(not_dhaka, add = TRUE, pch=20, cex=0.4, col = "red")

#This set of lines defines the object not_dhaka as the values of "points" that does no intersect with the map "dhaka", and then plots the bangladesh map map_bd and not_dhaka.


plot(dhaka, col = "gray", border = "blue", main = "Points Outside of Dhaka Map")
plot(not_dhaka, add = TRUE, pch=20, cex=0.4, col = "red")

#this set of lines plots the non-clipped points with the map of dhaka to demonstrate that the gDifference function was successful.

```
Area Calculations
```{r}

bd_utm = spTransform(map_bd, CRS("+proj=utm +datum=WGS84")) 
bd_utm$area = gArea(bd_utm, byid = TRUE) / 1000^2

head(bd_utm@data)

# Defines object bd_utm as map_bd projected into UTM, calculates the area as kilometers squared (function normally outputs meters, so denominator is required to convert units)

```




Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
