---
title: "Census Shenanigans"
output: html_notebook
author: Ethan Crookshank
---

Loading all of the packages that we need and using the API key to fetch census data
```{r}
# packages
library(tidyverse)
library(tidycensus)
library(sf)
library(ggplot2)

# settings for tidycensus
options(tigris_class = "sf")
options(tigris_use_cache = TRUE)
census_api_key("937eed644b7bc41f3ae13d18f51a459a4ac85100")

```


Alright, every single chunk after this does the Same. Exact. Thing.

First, we assign the 2019 data for average weekly hours worked in Maryland to the conveniently named mean_work_hours_2019 variable.

Next, we map it, because I was going to be ambitious and try to fulfill the "make a really good map with ggplots" challenge before my other classes clotheslined me, and I never took this line out. Post-hoc, it's probably good to make sure that it's actually plotting the data that you think should be in there.

We then convert to Mercator after plotting to prove that these ideas occurred to me at entirely different points in time.

Finally we write this data to a shapefile that we can load later in qgis or arcmap.
```{r}
mean_work_hours_2019 <- get_acs(geography = "tract", 
     variables = c("avg_work_hrs"="B23020_001"), 
     year = 2019,
     survey = "acs5",
     state =" MD", 
     
     geometry = TRUE, # download the shapefile with the data
     output = "wide") # need this



mean_work_hours_2019 %>% ggplot(aes(fill=mean_work_hours_2019$avg_work_hrsE, color=mean_work_hours_2019$avg_work_hrsE))+
  geom_sf() + 
  coord_sf(crs=26918) + # Web Mercator 3857, or UTM 18N for Maryland 26918
  scale_fill_viridis_c(option="magma", direction=-1) +
  scale_color_viridis_c(option="magma", direction=-1)

mean_work_hours_2019 <- st_transform(mean_work_hours_2019, 3857)


st_write(mean_work_hours_2019,  "MD_Mean_Work_Hours_2019.shp", update= T)

```
Same as above but for 2018
```{r}
mean_work_hours_2018 <- get_acs(geography = "tract", 
     variables = c("avg_work_hrs"="B23020_001"), 
     year = 2018,
     survey = "acs5",
     state =" MD", 
     
     geometry = TRUE, # download the shapefile with the data
     output = "wide") # need this



mean_work_hours_2018 %>% ggplot(aes(fill=mean_work_hours_2018$avg_work_hrsE, color=mean_work_hours_2018$avg_work_hrsE))+
  geom_sf() + 
  coord_sf(crs=26918) + # Web Mercator 3857, or UTM 18N for Maryland 26918
  scale_fill_viridis_c(option="magma", direction=-1) +
  scale_color_viridis_c(option="magma", direction=-1)

mean_work_hours_2018 <- st_transform(mean_work_hours_2018, 3857)


st_write(mean_work_hours_2018,  "MD_Mean_Work_Hours_2018.shp", update= T)
```
Next up is 2017
```{r}
mean_work_hours_2017 <- get_acs(geography = "tract", 
     variables = c("avg_work_hrs"="B23020_001"), 
     year = 2017,
     survey = "acs5",
     state =" MD", 
     
     geometry = TRUE, # download the shapefile with the data
     output = "wide") # need this



mean_work_hours_2017 %>% ggplot(aes(fill=mean_work_hours_2017$avg_work_hrsE, color=mean_work_hours_2017$avg_work_hrsE))+
  geom_sf() + 
  coord_sf(crs=26918) + # Web Mercator 3857, or UTM 18N for Maryland 26918
  scale_fill_viridis_c(option="magma", direction=-1) +
  scale_color_viridis_c(option="magma", direction=-1)

mean_work_hours_2017 <- st_transform(mean_work_hours_2017, 3857)


st_write(mean_work_hours_2017,  "MD_Mean_Work_Hours_2017.shp", update= T)
```

And now 2016
```{r}
mean_work_hours_2016 <- get_acs(geography = "tract", 
     variables = c("avg_work_hrs"="B23020_001"), 
     year = 2016,
     survey = "acs5",
     state =" MD", 
     
     geometry = TRUE, # download the shapefile with the data
     output = "wide") # need this



mean_work_hours_2016 %>% ggplot(aes(fill=mean_work_hours_2016$avg_work_hrsE, color=mean_work_hours_2016$avg_work_hrsE))+
  geom_sf() + 
  coord_sf(crs=26918) + # Web Mercator 3857, or UTM 18N for Maryland 26918
  scale_fill_viridis_c(option="magma", direction=-1) +
  scale_color_viridis_c(option="magma", direction=-1)

mean_work_hours_2016 <- st_transform(mean_work_hours_2016, 3857)


st_write(mean_work_hours_2016,  "MD_Mean_Work_Hours_2016.shp", update= T)
```




I'm not too good at math but I think 2015 comes next
```{r}
mean_work_hours_2015 <- get_acs(geography = "tract", 
     variables = c("avg_work_hrs"="B23020_001"), 
     year = 2015,
     survey = "acs5",
     state =" MD", 
     
     geometry = TRUE, # download the shapefile with the data
     output = "wide") # need this



mean_work_hours_2015 %>% ggplot(aes(fill=mean_work_hours_2015$avg_work_hrsE, color=mean_work_hours_2015$avg_work_hrsE))+
  geom_sf() + 
  coord_sf(crs=26918) + # Web Mercator 3857, or UTM 18N for Maryland 26918
  scale_fill_viridis_c(option="magma", direction=-1) +
  scale_color_viridis_c(option="magma", direction=-1)

mean_work_hours_2015 <- st_transform(mean_work_hours_2015, 3857)


st_write(mean_work_hours_2015,  "MD_Mean_Work_Hours_2015.shp", update= T)
```
On to 2014
```{r}
mean_work_hours_2014 <- get_acs(geography = "tract", 
     variables = c("avg_work_hrs"="B23020_001"), 
     year = 2014,
     survey = "acs5",
     state =" MD", 
     
     geometry = TRUE, # download the shapefile with the data
     output = "wide") # need this



mean_work_hours_2014 %>% ggplot(aes(fill=mean_work_hours_2014$avg_work_hrsE, color=mean_work_hours_2014$avg_work_hrsE))+
  geom_sf() + 
  coord_sf(crs=26918) + # Web Mercator 3857, or UTM 18N for Maryland 26918
  scale_fill_viridis_c(option="magma", direction=-1) +
  scale_color_viridis_c(option="magma", direction=-1)

mean_work_hours_2014 <- st_transform(mean_work_hours_2014, 3857)


st_write(mean_work_hours_2014,  "MD_Mean_Work_Hours_2014.shp", update= T)
```
2013, no fuss no muss
```{r}
mean_work_hours_2013 <- get_acs(geography = "tract", 
     variables = c("avg_work_hrs"="B23020_001"), 
     year = 2013,
     survey = "acs5",
     state =" MD", 
     
     geometry = TRUE, # download the shapefile with the data
     output = "wide") # need this



mean_work_hours_2013 %>% ggplot(aes(fill=mean_work_hours_2013$avg_work_hrsE, color=mean_work_hours_2013$avg_work_hrsE))+
  geom_sf() + 
  coord_sf(crs=26918) + # Web Mercator 3857, or UTM 18N for Maryland 26918
  scale_fill_viridis_c(option="magma", direction=-1) +
  scale_color_viridis_c(option="magma", direction=-1)

mean_work_hours_2013 <- st_transform(mean_work_hours_2013, 3857)


st_write(mean_work_hours_2013,  "MD_Mean_Work_Hours_2013.shp", update= T)
```
2012, when the world was supposed to end
```{r}
mean_work_hours_2012 <- get_acs(geography = "tract", 
     variables = c("avg_work_hrs"="B23020_001"), 
     year = 2012,
     survey = "acs5",
     state =" MD", 
     
     geometry = TRUE, # download the shapefile with the data
     output = "wide") # need this



mean_work_hours_2012 %>% ggplot(aes(fill=mean_work_hours_2012$avg_work_hrsE, color=mean_work_hours_2012$avg_work_hrsE))+
  geom_sf() + 
  coord_sf(crs=26918) + # Web Mercator 3857, or UTM 18N for Maryland 26918
  scale_fill_viridis_c(option="magma", direction=-1) +
  scale_color_viridis_c(option="magma", direction=-1)

mean_work_hours_2012 <- st_transform(mean_work_hours_2012, 3857)


st_write(mean_work_hours_2012,  "MD_Mean_Work_Hours_2012.shp", update= T)
```
I bet 2011 is next
```{r}
mean_work_hours_2011 <- get_acs(geography = "tract", 
     variables = c("avg_work_hrs"="B23020_001"), 
     year = 2011,
     survey = "acs5",
     state =" MD", 
     
     geometry = TRUE, # download the shapefile with the data
     output = "wide") # need this



mean_work_hours_2011 %>% ggplot(aes(fill=mean_work_hours_2011$avg_work_hrsE, color=mean_work_hours_2011$avg_work_hrsE))+
  geom_sf() + 
  coord_sf(crs=26918) + # Web Mercator 3857, or UTM 18N for Maryland 26918
  scale_fill_viridis_c(option="magma", direction=-1) +
  scale_color_viridis_c(option="magma", direction=-1)

mean_work_hours_2011 <- st_transform(mean_work_hours_2011, 3857)


st_write(mean_work_hours_2011,  "MD_Mean_Work_Hours_2011.shp", update= T)
```

And finally 2010, because that's as far back as this data was available for!
```{r}
mean_work_hours_2010 <- get_acs(geography = "tract", 
     variables = c("avg_work_hrs"="B23020_001"), 
     year = 2010,
     survey = "acs5",
     state =" MD", 
     
     geometry = TRUE, # download the shapefile with the data
     output = "wide") # need this





mean_work_hours_2010 %>% ggplot(aes(fill=mean_work_hours_2010$avg_work_hrsE, color=mean_work_hours_2010$avg_work_hrsE))+
  geom_sf() + 
  coord_sf(crs=26918) + # Web Mercator 3857, or UTM 18N for Maryland 26918
  scale_fill_viridis_c(option="magma", direction=-1) +
  scale_color_viridis_c(option="magma", direction=-1)

mean_work_hours_2010 <- st_transform(mean_work_hours_2010, 3857)


st_write(mean_work_hours_2010,  "MD_Mean_Work_Hours_2010.shp", update= T)

```
